---
title: "Introduction to `tmap`"
author: "Jason Flower, UC Santa Barbara"
date: "`r Sys.Date()`"
toc: true
number-sections: false
format: 
  html:
    self-contained: true
    code-tools: true
    df-print: paged
    fig-width: 8
    code-summary: "Show the code"
    code-overflow: wrap
---

# Introduction

This workshop will provide you with an introduction to making maps in R using the R package `tmap`. `tmap` is a relatively simple, but powerful, map making package. The code syntax will look somewhat familiar to `ggplot2` users. Although `ggplot2` can be used for map making, `tmap` is specifically designed for that purpose, and offers greater capabilities such as handling a wider variety of spatial data types and easy interactive mapping.

Basic knowledge of R and spatial data is assumed for this workshop.

We will cover:




*Resources*:

- The official `tmap` [website](https://r-tmap.github.io/tmap/){target="_blank"}, which includes many examples.
- [Elegant and informative maps with tmap](https://r-tmap.github.io/tmap-book/){target="_blank"}. An in-progress books written by the developers of `tmap`.
- The [Geocomputation with R: 9 Making Maps in R](https://r.geocompx.org/adv-map#static-maps){target="_blank"} book chapter. The entire books is available free online and is an excellent resource on map making and geospatial data in R in general.

## Prerequisites 

Apart from `tmap`, we will use the `terra` package for loading and manipulating spatial data. Both packages can be loaded using the following code

```{r}
#| eval: false

install.packages(c("terra", "tmap"))
```

We can now load the packages:

```{r}
library(terra)
library(tmap)
```

We will also load the data we will use.

```{r}
gbr_boundary <- vect("https://services-ap1.arcgis.com/8gXWSCxaJlFIfiTr/arcgis/rest/services/Great_Barrier_Reef_Marine_Park_Boundary_20/FeatureServer/replicafilescache/Great_Barrier_Reef_Marine_Park_Boundary_20_2474222838331219096.gpkg")


bathy <- ocenadatr::get_bathymetry()


```

